Algorithms for Hardware Caches

### 2017-05-11 paper.tex
Expand content, tweak things

@@

\usepackage{biblatex}
\addbibresource{paper.bib}

@@

% TODO: move this paragraph out of the introduction.
% Hardware caches are managed by hardware directly.  They are generally opaque to the
% operating system and other programs.  That is, software has no direct control over the
% contents of a hardware cache.

@@

% We will see that
Two algorithms solving the same problem with the same asymptotic complexity (in the same
\(\Theta(g(n))\)) may differ in performance by two orders of magnitude~\cite[p.~2]{afmh}.

% TODO: how and how much can we speed up our programs by considering their interactions
% with CPU caches?

% TODO: move this.
% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array to
% compute the sum of its elements.  Before exiting, it prints the CPU time spend summing the
% array.

@@

See \cref{fig:array-sum}.  What's going on here?

% \section{Basic Principles}
% \section{Basic Concepts}
\section{Key Terms}
\subsection{Spatial Locality}
Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
memory into the CPU cache) in \emph{blocks}.

@@

>>>>

% Use ISO 8601, like a reasonable person.  See <https://tex.stackexchange.com/a/152394>.
\usepackage[style=iso]{datetime2}

@@

% Keep using ISO 8601 consistently, like an even more reasonable person.  See
% <https://tex.stackexchange.com/q/231208>.
\usepackage[date=edtf, urldate=edtf]{biblatex}
\addbibresource{paper.bib}

@@

@@

% TODO: move this paragraph?
Hardware caches are managed by hardware directly.  They are generally opaque to the
operating system and other programs.  That is, software has no direct control over the
contents of a hardware cache.

% So what?  Why learn about hardware caches?  How much performance do I gain/lose
% depending on how *cache-friendly* my algorithm/code is?

% We will see that
Despite this, two algorithms solving the same problem with the same asymptotic complexity
(in the same \(\Theta(g(n))\)) may differ in performance by two orders of magnitude
because of different \emph{memory access patterns}~\cite{bigos}.  We will see an example
of this in section TODO.

In a nutshell, hardware caches are ubiquitous but the performance improvements they
provide are conditional.
\begin{comment}
   To use them effectively,
   % To obtain optimal performance,
   algorithms must be designed and implemented with the architecture
   % design, structure, manner of functioning, inner workings, properties
   of hardware caches in mind.
\end{comment}
Effective use of hardware caches requires knowledge about how they work.
% their architecture.
Algorithms must be designed and implemented observing this knowledge.
% their interactions with hardware caches.

% TODO: and *how* do I make effective use of hardware caches?

% \section{Basic Principles}
\section{Basic Concepts}
% \section{Key Terms}
% That sounds boring.

\subsection{Spatial Locality}
Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
memory into the CPU cache) in \emph{blocks}.

% TODO: move this.
% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array
% to compute the sum of its elements.  Before exiting, it prints the CPU time spend
% summing the array.

@@

% See \cref{fig:array-sum}.  What's going on here?

@@

\subsection{Cache Line} % or Cache Block
% On x86/x64, cache line is 64 bytes for many years now.
%    -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
